Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-14 talk/pepm2011/paper.tex
typo+tweak
some changes to sect. 4

The main insight to improve the code shown in the last section is that objects
in category 1 don't survive very long -- they are used only inside the loop and
nobody else in the program stores a reference to them. The idea for improving
the code is thus to analyze which objects fall in category 1 and may thus
not be allocated at all.

This is a process that is usually called \emph{escape analysis}. In this paper we will
perform escape analysis by using partial evaluation. The partial evalution is a
bit peculiar in that there are not actually any constant arguments to the trace,
but it is only used to optimized operations within a trace.

The partial evaluation works by walking the trace from beginning to end.
Whenever a \lstinline{new} operation is seen, the operation is removed and a static
object is constructed and associated with the variable that would have stored
the result of \lstinline{new}. The static object describes the shape of the
original object, \eg where the values that would be stored in the fields of the
allocated object come from, as well as the type of the object. Whenever the
optimizer sees a \lstinline{set} that writes into such an object, that shape
description is updated and the operation can be removed, which means that the
operation was done at partial evaluation time. When the optimizer encounters a
\lstinline{get} from such an object, the result is read from the shape
description, and the operation is also removed. Equivalently, a
\lstinline{guard_class} on a variable that has a shape description can be removed
as well, because the shape description stores the type and thus the outcome of
the type check the guard does is statically known.

In the example from last section, the following operations would produce two
static objects, and be completely removed from the optimized trace:

@@

The following operations on $p_{5}$ and $p_{6}$ could then be
optimized using that knowledge:

@@

The \lstinline{guard_class} operations can be removed, because the classes of $p_{5}$ and
$p_{6}$ are known to be \lstinline{BoxedInteger}. The \lstinline{get} operations can be removed
and $i_{7}$ and $i_{8}$ are just replaced by $i_{4}$ and -100. Thus the only
remaining operation in the optimized trace would be:

@@

The rest of the trace is optimized similarly.

So far we have only described what happens when static objects are used in
operations that read and write their fields and in guards. When the static
object is used in any other operation, it cannot stay static. For example, when
a static object is stored in a globally accessible place, the object needs to
actually be allocated, as it might live longer than one iteration of the loop
and because the partial evaluator looses track of it. This means that the static
object needs to be turned into a dynamic one, \ie lifted. This makes it
necessary to put operations into the residual code that actually allocate the
static object at runtime.

This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \lstinline{jump} operation
is hit. The arguments of the jump are at this point static objects. Before the
jump is emitted, they are \emph{lifted}. This means that the optimizer produces code
that allocates a new object of the right type and sets its fields to the field
values that the static object has (if the static object points to other static
objects, those need to be lifted as well, recursively) This means that instead of the jump,
the following operations are emitted:

@@

Note how the operations for creating these two instances have been moved down the
trace. It looks like for these operations we actually didn't win much, because
the objects are still allocated at the end. However, the optimization was still
worthwhile even in this case, because some operations that have been performed
on the lifted static objects have been removed (some \lstinline{get} operations
and \lstinline{guard_class} operations).

@@

In this paper, we used a approach based on online partial evaluation to
optimize allocations in the traces of a tracing JIT.  In this context a
simple approach to partial evaluation gives good results.  This is due
to the fact that the tracing JIT itself is responsible for all control
issues, which are usually the hardest part of partial evaluation: the
tracing JIT selects the parts of the program that are worthwhile to
optimize, and extracts linear paths through them, inlining functions as
necessary.  What is left to optimize is only those linear paths.

>>>>

The main insight to improve the code shown in the last section is that objects
in category 1 don't survive very long -- they are used only inside the loop and
nobody else in the program stores a reference to them. The idea for improving
the code is thus to analyze which objects fall in category 1 and thus do
not have to be allocated at all.

This is a process that is usually called \emph{escape analysis}. In this paper we will
perform escape analysis by using partial evaluation. The partial evalution is a
bit peculiar in that it receives no static input arguments for the trace,
but it is only used to optimize operations within a trace.

The partial evaluation works by traversing the trace from beginning to end.
Whenever a \lstinline{new} operation is seen, the operation is removed and a static
object is constructed and associated with the variable that would have stored
the result of \lstinline{new}. The static object describes the shape of the
original object, \eg where the values that would be stored in the fields of the
allocated object come from, as well as the type of the object. Whenever the
optimizer sees a \lstinline{set} that writes into such an object, that shape
description is updated and the operation can be removed, which means that the
operation was done at partial evaluation time. When the optimizer encounters a
\lstinline{get} from such an object, the result is read from the shape
description, and the operation is also removed. Equivalently, a
\lstinline{guard_class} on a variable that has a shape description can be removed
as well, because the shape description stores the type and thus the outcome of
the type check the guard does is statically known.

In the example from last section, the following operations in the upper half
of Fig.~\ref{fig:unopt-trace} would produce two
static objects, and would thus be completely removed from the optimized trace:

@@

The subsequent operations in Fig.~\ref{fig:unopt-trace},
 which use $p_{5}$ and $p_{6}$, could then be
optimized using that knowledge:

@@

First, the \lstinline{guard_class} operations can be removed, because the classes of $p_{5}$ and
$p_{6}$ are known to be \lstinline{BoxedInteger}. Second, the \lstinline{get} operations can be removed
and $i_{7}$ and $i_{8}$ are just replaced by $i_{4}$ and -100. Thus the only
remaining operation in the optimized trace would be:

@@

The rest of the trace from Fig.~\ref{fig:unopt-trace} is optimized similarly.

So far we have only described what happens when static objects are used in guards and in
operations that read and write fields. When the static
object is used in any other operation, it cannot remain static. For example, when
a static object is stored in a globally accessible place, the object has to
be allocated, as it might live longer than one iteration of the loop
and because the partial evaluator looses track of it. This means that the static
object needs to be turned into a dynamic one, \ie lifted. This makes it
necessary to put operations into the residual code that allocate the
static object at runtime.

This is what happens at the end of the trace in Fig.~\ref{fig:unopt-trace}, when the \lstinline{jump} operation
is hit. The arguments of the jump are at this point static objects. Before the
jump is emitted, they are \emph{lifted}. This means that the optimizer produces code
that allocates a new object of the right type and sets its fields to the field
values that the static object has (if the static object points to other static
objects, those need to be lifted as well, recursively). This means that instead of a simple jump,
the following operations are emitted:

@@

Observe how the operations for creating these two instances have been moved to later point in the
trace. 
At first sight, it may look like for these operations we didn't gain much, as
the objects are still allocated in the end. However, our optimizations were still
worthwhile, because some operations that have been performed
on the lifted static objects have been removed (some \lstinline{get} operations
and \lstinline{guard_class} operations).

@@

In this paper, we used an approach based on online partial evaluation to
optimize away allocations in the traces of a tracing JIT.  In this context a
simple approach to partial evaluation gives good results.  This is due
to the fact that the tracing JIT itself is responsible for all control
issues, which are usually the hardest part of partial evaluation: the
tracing JIT selects the parts of the program that are worthwhile to
optimize, and extracts linear paths through them, inlining functions as
necessary.  What is left to optimize is only those linear paths.
